---
title: "BIOS 1 - Homework 2"
output:
  html_document: 
    toc: yes
date: "2023-05-03"
author: "Levi Waldron"
editor_options:
  markdown:
    wrap: 72
---

# Source code and data for this assignment

*Note*: The code used to create this document is [here](https://github.com/lwaldron/bios1/blob/main/vignettes/Homework2.Rmd).

In this homework we will again use the NHEFS dataset. Take a look at the codebook linked to from
<https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/>.

This time we will test the hypothesis that quitting smoking between 1971 and 1981 
(variable `qsmk`) reduces the probability of death by 2001 (variable `death`). 

# Data import and recoding

Download the dataset, import it into R, and recode variables as instructed in
[Homework 1](https://lwaldron.github.io/bios1/articles/Homework1.html).

```{r, cache=TRUE, eval = FALSE}
library(readr)
nhefs <- read_csv("https://cdn1.sph.harvard.edu/wp-content/uploads/sites/1268/1268/20/nhefs.csv")
```

# Question 1 (2 points): filter any censored or missing data

Remove all participants that were lost to follow-up or "censored", whose value 
of the `cens` variable is 0. You would ordinarily want to perform a sensitivity 
analysis and consider the possibility of bias from informative censoring, but 
you can ignore that possibility for the purposes of this assignment.

Also remove observations that are missing any of the following variables:

-   `death`
-   `cens`
-   `active`
-   `age`
-   `education`
-   `exercise`
-   `qsmk`
-   `race`
-   `sex`
-   `smokeintensity`
-   `smkyrs`
-   `wt82_71`

# Question 2 (4 points): Table 1

Create a table of descriptive statistics like in HW1, but this time stratify it
by the exposure variable `qsmk`


# Question 2 (4 points): fit crude and adjusted logistic regression models

Fit simple and adjusted multiple logistic regression models for the probability of death
with quitting smoking as the exposure. In the multivariate model, adjust for sex, race, age, education, and 
exercise as hypothesized confounders. Use a B-spline from the `splines` package to correct for age, ie:

$$
death \sim qsmk + sex + race + bs(age) + education + exercise 
$$

For the purpose of this question, assume that conditional exchangeability holds when 
controlling for these confounders. 

**Interpret the coefficient of `qsmk` in the crude model and in the adjusted model. 
Compare them.**

# Question 3 (3 points): model diagnostics

Create a binned residual plot using the `binnedplot` function from the `arm` 
library. Interpret the plot.

# Question 4 (7 points): Inverse probability weighting

a (2 points)) Create a model to calculate the propensity of exposure given the values of the 
hypothesized confounders (propensity score). This is your "denominator" model of 
Lecture 11, Slide 15.  ie,

$$
\textbf{denom.fit}:qsmk \sim sex + race + bs(age) + education + exercise 
$$

Also fit your "numerator model",
$$
\textbf{num.fit}: qsmk \sim 1
$$


b (2 points)) Calculate IPTW weights using this model. Hint - you can use code like this:

```{r, eval = FALSE}
pdenom <- predict(denom.fit, type = "response")
pnum <- predict(num.fit, type = "response")
nhefs$iptw <- ifelse(nhefs$qsmk == 0, ((1-pn.qsmk)/(1-pd.qsmk)),
                     (pn.qsmk/pd.qsmk))
```

c (2 points)) Use the `geeglm` function from the `geepack` library to use these weights to
estimate the causal effect of quitting smoking on death. 

Notes:
- the `id` argument to `geepack` is arbitrary in this case because there is no grouping or
hierarchical structure of participants. You can use the `seqn` variable or `1:nrow(nhefs)`.
- The `corstr` argument is also arbitrary for the same reason; use `corstr="unstructured"`
- Remember that when using these weights, *do not* include confounders in the model.


d (1 point)) Compare the coefficient and standard error for quitting smoking to Question 2, 
and comment. Recall the discussion in lecture about `geepack` standard errors being 
too conservative.

