---
title: "Lab 6: Image analysis to predict malignant breast cancer logistic regression"
author: "Biostatistics 1"
output: 
  html_document: 
    toc: yes
---

# Learning Objectives

* Identify and correct a problematic data file
* perform exploratory data analysis with a binary variable
* perform logistic regression in R using the `glm()` function
* identify collinearity among many numeric variables
* interpret logistic regression coefficients in a predictive setting
* use a logistic regression model to predict probabilities of a binary outcome
* use `tidyr` to pivot from a wide format to a long format dataframe
* use ggplot2's `facet_wrap()` to create multiple plots
* overlay a line plot of predicted probabilities over an xy plot of data points


# Materials

* RStudio or any R environment
* Dataset: `breast cancer.csv` ([download](https://raw.githubusercontent.com/lwaldron/bios1/main/vignettes/breast%20cancer.csv)).
This dataset contains an outcome variable "diagnosis" with values B (benign) and M (malignant), and 
a number of cellular pathologic features representing the properties of cells as calculated by an
image analysis algorithm. Each row corresponds to a patient who underwent a tissue biopsy. Here is [more information about the dataset](http://archive.ics.uci.edu/ml/datasets/Breast+Cancer+Wisconsin+%28Diagnostic%29) if you are interested.
* R packages: readr, dplyr, ggplot2, tidyr, possibly ComplexHeatmap


# 1. Importing the Dataset

a. Import the dataset (`breast cancer.csv`) into R, using `readr`. What is the 
problem with the 33rd column causing a warning?  Fix this problem in the data 
file, then try again. Import `id` as a character, and `diagnosis` as a factor 
with values B and M.
b. Recode the "diagnosis" column to the more informative values "benign" and "malignant", 
with reference level "benign"

# 2. Exploratory Data Analysis (EDA)

For each part of this question, write out something you notice from the data
exploration, such as variable type, if and where there are missing values. 
There are no right/wrong answers for this, it is just to get practice interpreting EDA. 

a. Check the dimensions of the dataset using the dim() function

b. Preview the first few rows of the dataset using the head() function

c. Use `summary()` to summarize the dataset. 

d. Identify collinear variables.

The following command, run on a wide dataframe, will calculate 1 minus the pairwise 
Pearson correlation between each pair of numeric variables in the dataset. This
represents a distance matrix between variables, where "0" means two variables 
that are identical or perfectly anti-correlated, and "1" means two variables have zero
correlation. We use 1 minus the absolute value of the correlations so that it 
becomes a distance measure instead of a correlation measure. 

Plot this distance matrix to identify the most highly correlated or anticorrelated variables. 
A few ways you could do this is (if your distance matrix is called `d`) are as
follows - try them and choose your favorite. You may want to increase the size 
of the figure output again.

```{r, eval = FALSE}
plot(hclust(as.dist(d)))
```

or:
```{r, eval=FALSE}
ComplexHeatmap::Heatmap(d)
```

or:
```{r, eval=FALSE}
ComplexHeatmap::pheatmap(d)
```

e. Create a box plot for each column except for `id`.  Each boxplot should have 
two boxes, one for benign and one for malignant, allowing visual comparison of the
distribution of that variable for benign and malignant specimens. Do any variables clearly
have an association with breast cancer diagnosis?

_Hints for e:_

1. Create a new dataset without the `id` variable, then use the `tidyr::pivot_longer`
function to create a "long" dataframe with 3 columns: 1. diagnosis, 2. the name of 
the column from the "wide" dataframe, and 3. a column containing all the numeric values.
The first example from the `pivot_longer` help page ("Simplest case where column names 
are character data") is exactly analogous to what you need to do.
2. With this "long" dataset you can add a `facet_wrap()` command to your ggplot 
to create a box plot for each variable in a grid. Use the `scales` argument
to use "free" scales so that each box plot can have a different scale, so that you 
can read each box plot.
3. Use the knitr chunk options `fig.width=10` and `fig.height=10` to increase the
size of the figure to make theaxis labels readable.

# 3. Building a Logistic Regression Model

a. Fit a univariate logistic regression model using `area_mean` as the predictor and 
`diagnosis` as the outcome, using the `glm()` function

b. Print the summary of the model using the summary() function

c. Interpret the coefficients of the logistic regression model

# 4. Making predictions

a. Create a xy plot with `area_mean` of the x axis and `diagnosis` on the y axis.
Use `geom_jitter(width = 0)` to create some spread on the y-axis so you can see where 
the points are without changing the x values.

b. Now add to this plot the predicted probabilities for each observed temperature, using
the `predict` function with argument `type="response"`.

_Hint:_ First add a column of predicted probabilities to the `breast_cancer`
dataframe. You will have to add 1 to these probabilities to make them on the same
scale as the data points. Then you can add a `geom_line()` to the previous plot, 
with its own `aes()`, to add the line. 

_Suggestion:_ Change the line color and increase its width to make it more visible.

c. What do you notice about the data points where the predicted probabilities are 0, 0.5, and 1?

d. What are the predicted probabilities of a malignant diagnosis when `area_mean` is 300, 500,
700, 900, and 1100?
