---
title: "Standardization simplified"
author: "Levi Waldron"
date: "`r Sys.Date()`"
output: 
  html_document: 
    toc: yes
    number_sections: yes
---

# Simulate some data

Here we simulate a sample of 20 participants with observations on:
* _sex_ (M or F)
* _height_ in cm (from a random normal distribution with mean 180cm and standard 
deviation 10 cm), plus 5cm for males

```{r}
set.seed(1)
df <- data.frame(
  sex = c(rep("M", 10), rep("F", 10)),
  group = sample(c(rep("trt", 10), rep("placebo", 10)), size = 20, replace = FALSE)
)
df$height = rnorm(20, mean = 180, sd = 10) + ifelse(df$sex == "M", 5, 0)
df$weight <-
  df$height * 0.2 + ifelse(df$sex == "M", 5, 0) - ifelse(df$group == "trt", 2, 0) + rnorm(20)
```

# Estimate the effect of treatment using regression

```{r}
mod <- lm(weight ~ height + sex + group, data = df)
summary(mod)
```
# Estimate the effect of treatment using standardization

Need to create new datasets where everyone was treated, or everyone received
placebo.

```{r}
newdatatrt <- df
newdatatrt$group <- "trt" #everyone treated
newdataplacebo <- df
newdataplacebo$group <- "placebo" #everyone placebo
```

What would the average weight be if everyone received treatment?

```{r}
(trtmean <- mean(predict(mod, newdata = newdatatrt)))
```

What would the average weight be if everyone received placebo?

```{r}
(placebomean <- mean(predict(mod, newdata = newdataplacebo)))
```

What is the causal effect of treatment compared to placebo in this sample?

```{r}
trtmean - placebomean
```

How does this compare to the estimate from the regression table? It's THE SAME.
But if there were any interactions with treatment, we would not find the standardized
mean treatment effect anywhere in the table of regression coefficients.

Example: if the treatment had a different effect on males than on females, the
regression table would only show the effect for the reference sex. You could add
the coefficient of the interaction term with sex to get the effect of treatment on
the non-reference sex. The standardized effect would be the average effect on
the composition of males and females found in this sample. The average treatment
effect could be recalculated for any other composition of males and females. 

# Why standardization then?

1. If there were interactions with treatment, then the effect of treatment from the
regression output would only be for the reference group, it would not be an average
treatment effect for the sample.

2. We can standardize to different samples or to the population, where the distribution
of covariates is different than in our sample, even in the presence of interactions.

3. Standardization is easily amenable to bootstrap simulation to estimate confidence intervals.
