---
title: "Tutorial 5"
output:
  html_document: 
    toc: yes
date: "`r Sys.Date()`"
author: "Lisa Hitch"
editor_options:
  markdown:
    wrap: 72
---

The `eval = FALSE` in this chunk prevents code from being evaluated on knit:

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Importing data

This tutorial is supposed to teach you how to load your dataset and "clean" your data (i.e. rename variables or levels of variables, remove irrelevant variables, code missing data etc.)

When you have a dataset, click on File, then "Import Dataset" and then "From Text (Readr)". Choose your file and it should show you a preview of your dataset. On the top, you see your variable names and next to it is a little arrow. Click that and you can choose different variable types, for example "character" for text data, "double" for numeric data, or "factor" for variables with 0 and 1. You can also skip variables here if you do not want them imported. Before you click "Import", copy the code on the bottom right and paste it in a code chunk in your R Markdown document.

Import the dataset "diabetes" as described above and choose the correct variable type for each variable. Paste the code that you copy from the readr here:

```{r}

```

### Troubleshooting

There is one column being imported as a character variable, when it should be
numeric. Why? Fix it in the data file, then import again.


## Recode impossible variable values as missing
  
Part of data cleaning is sometimes recoding impossible variable values as missing.
For example, blood pressure values of "0" or "-99" or "999" are impossible and represent
missing values. Use dplyr's `case_when` or `case_match` functions to recode any 
such impossible values as `NA`.

```{r}

```


## Subsetting 

Create a subset, for example only participants who had exactly 1 pregnancy.

```{r}
filter(diabetes, Pregnancies == 1) |>
  pull(Pregnancies) |>
  table()
```
## Keep/Remove/Rename variables manually

You can also add or remove variables after you imported your dataset with the 
`dplyr::select` function (see code below). The `dplyr::rename()` function can be 
used to change the name of your variable. It always follows this structure: 
* `old name = new name`

```{r}
library(dplyr)
diabetes1 <- select(diabetes, c(BMI, Glucose, Insulin)) #this only keeps these three variables
diabetes1

diabetes2 <- select(diabetes, -Pregnancies) #this removes the Pregnancy variable
diabetes2

diabetes3 <- rename(diabetes, BodyMassIndex = BMI) #this renames the old variable "BMI" to "BodyMassIndex"
diabetes3


```

### Exercise 2

Create a diabetes dataset that only includes Outcome, Insulin, Glucose, Age, and BMI and rename Glucose to "blood_glucose"

```{r}

```

## 3. Change variable type

In same cases you want to change the variable type. For example, your variable 
has 0 and 1, but since they are not numeric but stand for two different levels 
of your variable (e.g. female and male) you want to convert it to a factor. 

```{r}
#convert any variable to a string or character variable
diabetes4 <- mutate(diabetes, Outcome=as.character(Outcome))
class(diabetes4$Outcome)

#convert any variable to a numeric variable
diabetes4 <- mutate(diabetes, Outcome=as.numeric(Outcome))
class(diabetes4$Outcome)

#convert any variable to a factor variable
diabetes4 <- mutate(diabetes, Outcome=factor(Outcome))
class(diabetes4$Outcome)
```

## 4. Delete `NA` or missing data

Note that you should never delete missing data before creating your Table 1
(ie missingness should be included in your table of descriptive statistics). 
In your analysis there are several ways to deal with missingness, including:
* multiple imputation
* for factors, treating missing values as a distinct factor level
* removal of observations with missing values for any of the variables of interest
(ie outcome or covariates in a regression model)

You can use the following code to any observations with any missing data ("NA"). 
This should _only_ be done if you have already selected only columns of interest,
otherwise you might remove rows that were only missing irrelevant variables.

```{r}
diabetes5 <- na.omit(diabetes) #this deletes all rows with missing data
```

Note, it is _NOT_ recommended to use integer values to represent missingness. In
R that is what `NA` is for. The above command will only remove missing values 
that are coded as NA.


### Exercise 3:
Create a dataset from the diabetes dataset with only Elderly adults who are older than 60 years.

```{r}

```

## 6. Rename levels of your variable
We may want to re-name the levels of a variable. So first, I would check the levels and the level order with the levels() function.

```{r}
levels(diabetes6$Outcome)
```
So here I see that the Outcome variable has two levels: 0 and 1. All datasets should have a codebook where it states what the 0 and 1 stand for. Here, the 0 stands for not having a diabetes diagnosis and 1 stands for having a diabetes diagnosis. To make it easier, I can rename the levels with the following code:

```{r}
levels(diabetes6$Outcome) <- c("no diagnosis", "diagnosis")
diabetes6

diabetes8 <- diabetes %>% mutate(Outcome = case_match(Outcome, 
"0" ~ "No diagnosis",
"1" ~ "diagnosis"))
diabetes8

```

## Exercise 4:
Now that you have some experience cleaning your data, try the following exercise:
Create a dataset from the diabetes dataset with the following characteristics:

Drop the variable DiabetesPedigreeFunction (keep all other variables)
Age between 20-60
Outcome as a factor and 0 = no diabetes and 1 = diabetes
Insulin, Skin Thickness, and Blood Pressure > 0
BMI between 15 and 40

```{r}

```


